A quantitative framework for investigating the reliability of network construction

### 2018-01-17 manuscript/CI_interactions.tex
some small updates

  Current descriptions of ecological networks will typically assume that specific interactions occur wherever and whenever the community is observed. This runs in stark contrast with the widely recognized stochastic nature of ecological communities. When accurately resolved, community composition, species abundances and link structure have typically been found to vary in space and time. On top of this, finite sampling will generate variation in what links are observed and what not. Despite these challenges, we still lack the quantitative methodology to deal with the uncertainty generated by “true” spatiotemporal variation in ecological interactions and by sampling. In this study, we develop analytical tools to capture the uncertainty in the estimation of pairwise interactions. To define the problem, we first discuss the pervasive implications of uncertainty in the estimation of pairwise interactions on the properties of ecological networks. We then define a framework which identifies and dissects the different contributions to uncertainty of interaction. We illustrate this framework with a simple quantitative example using the most extensively sampled network to date. Next, we use the framework to provide analytical criteria for improving the sampling of rare interactions. Finally, we offer tangible recommendations for improved descriptors of ecological interactions. Through these efforts, we demonstrate both the utility of our approach and the importance of acknowledging the uncertainty inherent in network studies.

  % Can be up to 200 words

@@

    Representation of a species assemblage as a network offers a convenient summary of how the community is constructed as it simultaneously accounts for species composition and their interactions. A tabulation of the nodes (species) and of their relative abundances forms the basis for traditional metrics of community composition. On top of this information, interactions are represented by links between nodes so that networks provide additional, higher-order information on community structure. Empirical descriptions of ecological networks are, however, usually considered to be static representations of the communities and interactions they describe. That is, whether the network is assembled based on aggregated data, a single intensive ``snapshot" sample, or expert knowledge, interactions are assumed to occur deterministically wherever and whenever the community is observed~\citep{Olesen2011a}.

    The assumption of static communities contrasts significantly with the widely recognized stochasticity of ecological communities \citep{Gotelli2000}. Community composition and species abundances vary from site to site~\citep{Baiser2012} and over time within a site~\citep{Olesen2011a}. Likewise, interactions vary over space~\citep{Kitching1987,Baiser2012}, time~\citep{Kitching1987,Olesen2011a}, and between individuals of a given species~\citep{Pires2011a,Fodrie2015,Novak2015}. Moreover, variability in community composition and interactions may or may not be closely related. The removal of a species from a site will obviously also remove its interactions, but conversely, the co-occurrence of potentially interacting species does not in itself guarantee that they will interact at a given place and time. Interactions can be lost if the interaction partners remain present but are separated in time or are too rare to detect each other~\citep{Tylianakis2010}. Interactions can also fail to occur because of environmental contingencies \citep{Poisot2015}, or through changes to individual preferences~\citep{Fodrie2015}.

    Beyond ``true" variation in network structure, several researchers have pointed to the importance of sampling intensity for the assessment of network structure (e.g.,~\citealp{Martinez1999,Bluthgen2006,Bluthgen2007}). An assessment of the accumulation of interactions with increasing sampling effort suggests that it is even more challenging to document interactions than species richness \citep{Poisot2012}. As a result, it has been proposed that interactions should be described probabilistically and network metrics computed accordingly \citep{Poisot2016}. Yet, to dat, we lack the quantitative methodology to deal with the uncertainty generated by spatiotemporal variation in ecological interactions and by sampling. Moreover, even in extremely well-sampled networks uneven sampling across species (or pairs of species) can lead to the erroneous inference that some species do not interact because they co-occur rarely or have not yet been observed together - even if they do interact when they do co-occur (see Box 1 for an example). Nearly all network studies will thus neglect some interactions, necessitating an approach that acknowledges this uncertainty.

    In this study, we develop analytical tools to capture the uncertainty in the estimation of pairwise interactions. We focus on binary interactions as a first step, but the framework could be expanded to deal with interaction frequencies and strength. To define the problem, we first discuss the pervasive implications of uncertainty in the estimation of pairwise interactions on the properties of ecological networks. We then define a framework which identifies and dissects the different contributions to uncertainty of interaction. We illustrate this framework with a simple quantitative example using the most extensively sampled network to date (Box 1). Next, we use the framework to provide analytical criteria for improving the sampling of rare interactions. Finally, we offer tangible recommendations for improved descriptors of ecological interactions. Through these efforts, we demonstrate both the utility of our approach and the importance of acknowledging the uncertainty inherent in network studies.

@@

  To define the problem associated with quantifying ecological interaction networks, we will start from the perspective of an empirical community ecologist faced with the task of describing a previously unknown interaction network. This ecologist will be interested in generating a description of the species/nodes present, and the links between them~\citep{Roslin2016}.  Importantly, the information sought is conveyed by both the presence and \emph{absence} of links. The two types of information are not, however, equally certain. An observed link will always remain an observed link, but there are multiple reasons why a given link may not be observed. Thus, the detection of any interaction is a stochastic process. We define three nested levels of uncertainty contributing to this stochasticity:

@@

    First, and most fundamentally, we do not know whether or not a pair of species have the appropriate characteristics (or traits) to interact. We define the probability of an interaction $L$ given those characteristics $\mathbf{T}$ as $P(L | \mathbf{T})=\lambda$. Obviously, if $k$ (the number of observed interactions) is 0, it is possible that the two species could not interact even if there were no external constraints (e.g., temporal or environmental separation) preventing the interaction from occurring. As a simple example, a prey species may be too large to be consumed by a particular predator. In such cases, $\lambda$ would take a value of 0. 
    % [[The previous sentence doesn't follow, k hasn't been defined and we don't know that it's zero. Also, are we assuming that these interactions have been observed over a large area/time like the Kopelke data? Otherwise couldn't something about that particular site prevent species interacting even if they otherwise could?]] 
    % [[With defining k I think this is fine... site uncertainty will come in later, and it's not necessary to assume anything about the sampling design to get from not observing two species to interact to suspecting that they don't (degree of confidence depending on the species involved). I've toned down "very likely" to "possible" though - does that help?]]
    Nevertheless, it is also possible that the interaction is a rare phenomenon with $\lambda>0$ that has not yet been documented. This source of uncertainty is the one documented by trait-matching models~\citep{Bartomeus2016}. It arises because every model is imperfect and lacks information (i.e. about traits) defining the constraints on the interaction. In other words, with sufficient sampling and all information accessible, this interaction probability $\lambda$ should either tend to 0 or to 1 and the uncertainty vanishes. 
    % [[For species which actually co occur (?)]]
    % [[I think that gets covered under "sufficient sampling"]]

@@

    Even when an interaction is feasible, i.e. $L=1$, it may not occur at a given location or moment in time because of local constraints such as inclement weather or the lack of suitable habitat. We define the realisation of the interaction process with the variable $X$, given that the interaction is feasible, as a stochastic process with associated probability $P(X|L)=\chi$. This phenomenon of interaction contingencies is usually not considered in network studies, but there is a rich literature in community ecology about the contingencies of interactions. Phenological matching~\citep{MillerRushing2010,Gezon2016}, species preferences~\citep{Pires2011,Novak2015,Coux2016}, and fear effects of other species~\citep{} are just some of the factors contributing to variation in the frequency of interactions between a given pair of species.
    % We imagine for instance that an interaction between a gall and a parasitoid might not be recorded at a location because it was the wrong time of the year and the gall had not yet formed (phenological constraints), a late frost killed the larvae (an abiotic environmental constraint), another parasitoid species competitively excluded the parasitoid species of interest (a biotic environmental constraint) or simply the parasitoid is too rare to interact with all of its viable hosts (an abundance constraint).

@@

    Lastly, measurement errors are a pervasive source of uncertainty in the observation of ecological processes. Given that an interaction with the variable $D$, given an interaction is feasible and occurs under the local conditions, we may define the detection of an interaction, $D$, as a stochastic process with the associated probability $P(D|X,L)=\delta$. Detection failure could happen for several reasons, for instance if the rearing of a parasitoid fails with inappropriate lab conditions, because of species mis-identification, or because the interaction is very rare (see~\citet{Wirta2014} for examples of some of these difficulties and partial solutions to them). Some sources of detection error could be minimised with appropriate sampling effort ($\delta$ will converge to one with increasing number of samples), but other sources are often difficult to reduce (e.g. the occurrence of cryptic species might require molecular analysis for appropriate taxonomic identification as in~\citet{Wirta2014,Frost2016}).

@@

    The considerations above raise a major challenge: when faced with empirical data, how may we infer whether unobserved interactions went undetected due to sampling, or whether they truly do not occur. How then may we refine our sampling approaches to reduce uncertainties, and to gain insights into the impact of multiple processes on field observations? Importantly, some sources of uncertainty can be minimised with appropriate sampling design and efforts while other sources are difficult or impossible to reduce since they are generated by chance variation created by the very process in which we are interested.

@@

          As a fourth and final approach to obtain the prior distribution of $\lambda$, we envisage using the outcome of a trait-matching model, provided such a model has been documented using external data and relevant traits are available. 
          In such as case, the prior distribution would follow the function $P(\lambda|\mathbf{T})=f(\mathbf{T})$ based on a set of traits for both species $\mathbf{T}$. There are several techniques available to perform this inference of interaction probability, some of which are Bayesian, and we refer to ~\citep{Bartomeus2016} and \citep{Weinstein2017} for recent reviews about this topic. Note that in this case the prior might not be beta distributed and numerical methods might be required to compute the posterior distribution.

>>>>

  Descriptions of ecological networks typically assume that the same interspecific interactions occur wherever and whenever a community is observed. This contrasts with the known stochasticity of ecological communities: community composition, species abundances, and link structure all vary in space and time. On top of this, finite sampling will generate variation in the set of interactions that are actually observed. Despite recognising these challenges, we lack the quantitative methodology to deal with the uncertainty generated by “true” spatiotemporal variation in ecological interactions or by incomplete sampling. Here we develop analytical tools to capture the uncertainty in the estimation of pairwise interactions. To define the problem, we identify the different contributions to the uncertainty of an interaction and discuss the implications of this uncertainty in the estimation of network properties. We then develop a framework to explicitly quantify the uncertainty around each interaction. We illustrate this framework using the most extensively sampled network to date, demonstrating the universal uncertainty in network studies. Finally, we offer tangible recommendations for improving descriptions of ecological networks and for better sampling of rare interactions. Through these efforts, we demonstrate both the utility of our approach and the importance of acknowledging the uncertainty inherent in network studies.

  % Can be up to 200 words. This is 200 exactly.
  % Word count can be up to 7500, we're at about 5700 without the boxes. Should be grand.

@@

    Representating an assemblage of species as a network offers a convenient summary of how the community is constructed as it simultaneously accounts for species composition and interactions between species. A tabulation of the nodes (species) and their relative abundances forms the basis for traditional metrics of community composition. To move from these simpler metrics to a network framework, the tabulation of nodes is combined with interactions (links between nodes) so that networks provide additional, higher-order information on community structure. While this additional information is useful (as, for example, interactions can affect changes in species abundances over time), empirical descriptions of ecological networks are still limited because they are usually considered to be static representations of the communities and interactions they describe. That is, whether the network is assembled based on aggregated data, a single intensive ``snapshot" sample, or expert knowledge, interactions are assumed to occur deterministically wherever and whenever the community is observed~\citep{Olesen2011a}.

    The assumption of static communities contrasts significantly with the widely recognized stochasticity of ecological communities~\citep{Gotelli2000}. Community composition and species abundances vary from site to site~\citep{Baiser2012} and over time within a site~\citep{Olesen2011a}. Likewise, interactions vary over space~\citep{Kitching1987,Baiser2012}, time~\citep{Kitching1987,Olesen2011a}, and between individuals of a given species~\citep{Pires2011a,Fodrie2015,Novak2015}. Moreover, variability in community composition and interactions may or may not be closely related. The removal of a species from a site will obviously also remove its interactions but, conversely, the co-occurrence of potentially interacting species does not in itself guarantee that they will interact at a given place and time. Interactions can be lost if the interaction partners remain present but are separated in time or are too rare to detect each other~\citep{Tylianakis2010}. Interactions can also fail to occur because of environmental contingencies~\citep{Poisot2015}, or through changes to individual preferences~\citep{Fodrie2015}.

    Beyond ``true" variation in network structure, several researchers have pointed to the importance of sampling intensity for the assessment of network structure (e.g.,~\citealp{Martinez1999,Bluthgen2006,Bluthgen2007}). An assessment of the accumulation of interactions with increasing sampling effort suggests that it is even more challenging to document interactions than species~\citep{Poisot2012}. As a result, it has been proposed that interactions should be described probabilistically and network metrics computed accordingly~\citep{Poisot2016}. Yet, to date, we lack the quantitative methodology to deal with the uncertainty generated by spatiotemporal variation in ecological interactions and by sampling. Even in extremely well-sampled networks, uneven sampling across species (or pairs of species) can lead to the erroneous inference that some species do not interact because they co-occur rarely or have not yet been observed together - even if they do interact when they do co-occur (see Box 1 for an example). Nearly all network studies will thus neglect some interactions, necessitating an approach that acknowledges this uncertainty.

    In this study, we develop analytical tools to capture the uncertainty in the estimation of pairwise interactions. We focus on binary interactions as a first step, but the framework could be expanded to deal with interaction frequencies and strength. To define the problem, we first identify the different contributions to the uncertainty of an interaction and discuss the implications of each source of uncertainty for the properties of ecological networks. Next, we develop an analystical framework to quantify the uncertainty around interactions in an empirical web. We illustrate this framework using the most extensively sampled network to date (Box 1).  Finally, we offer tangible recommendations for improved descriptors of ecological interactions [[double-check that we do this]]. Through these efforts, we demonstrate both the utility of our approach and the importance of acknowledging the uncertainty inherent in network studies.

@@

  To define the problems associated with quantifying ecological interaction networks, we will start from the perspective of an empirical community ecologist faced with the task of describing a previously unknown interaction network. This ecologist will be interested in generating a description of the species/nodes present, and the links between them~\citep{Roslin2016}.  Importantly, the information sought is conveyed by both the presence and \emph{absence} of links. The two types of information are not, however, equally certain. An observed link will always remain an observed link, but there are multiple reasons why a given link may not be observed. Thus, the detection of any interaction is a stochastic process. We define three nested levels of uncertainty contributing to this stochasticity:

@@

    First, and most fundamentally, we do not know whether or not a pair of species have the appropriate characteristics (or traits) to interact. We define the probability of an interaction $L$ given those characteristics $\mathbf{T}$ as $P(L | \mathbf{T})=\lambda$. Obviously, if $k$ (the number of observed interactions) is 0, it is possible that the two species could not interact even if there were no external constraints (e.g., temporal or environmental separation) preventing the interaction from occurring. As a simple example, a prey species may be too large to be consumed by a particular predator. In such cases, $\lambda$ would take a value of 0. 
    % [[The previous sentence doesn't follow, k hasn't been defined and we don't know that it's zero. Also, are we assuming that these interactions have been observed over a large area/time like the Kopelke data? Otherwise couldn't something about that particular site prevent species interacting even if they otherwise could?]] 
    % [[With defining k I think this is fine... site uncertainty will come in later, and it's not necessary to assume anything about the sampling design to get from not observing two species to interact to suspecting that they don't (degree of confidence depending on the species involved). I've toned down "very likely" to "possible" though - does that help?]]
    Nevertheless, it is also possible that the interaction is a rare phenomenon with $\lambda>0$ that has not yet been documented. This source of uncertainty is the one documented by trait-matching models~\citep{Bartomeus2016}. It arises because every model is imperfect and lacks information (i.e. about traits) that could be used to define constraints on the interaction. In other words, with sufficient sampling and all information accessible, this interaction probability $\lambda$ should either tend to 0 or to 1 and the uncertainty should vanish. 
    % [[For species which actually co occur (?)]]
    % [[I think that gets covered under "sufficient sampling"]]

@@

    Even when an interaction is feasible, i.e. $L=1$, it may not occur at a given location or moment in time because of local constraints such as inclement weather or the lack of suitable habitat. We define the realisation of the interaction process with the variable $X$, given that the interaction is feasible, as a stochastic process with associated probability $P(X|L)=\chi$. This phenomenon of interaction contingencies is usually not considered in network studies, but there is a rich literature in community ecology about the contingencies of interactions. Phenological matching~\citep{MillerRushing2010,Gezon2016}, species preferences~\citep{Pires2011,Novak2015,Coux2016}, and fear effects of other species~\citep{Luttbeg2005,Wirsing2007} are just some of the factors contributing to variation in the frequency of interactions between a given pair of species.
    % We imagine for instance that an interaction between a gall and a parasitoid might not be recorded at a location because it was the wrong time of the year and the gall had not yet formed (phenological constraints), a late frost killed the larvae (an abiotic environmental constraint), another parasitoid species competitively excluded the parasitoid species of interest (a biotic environmental constraint) or simply the parasitoid is too rare to interact with all of its viable hosts (an abundance constraint).

@@

    Lastly, measurement errors are a pervasive source of uncertainty in the observation of ecological processes. Given that an interaction with the variable $D$, given an interaction is feasible and occurs under the local conditions, we may define the detection of an interaction, $D$, as a stochastic process with the associated probability $P(D|X,L)=\delta$. Detection failure could happen for several reasons, for instance if the rearing of a parasitoid fails with inappropriate lab conditions, because of species mis-identification, or because the interaction is very rare (see~\citet{Wirta2014} for examples of some of these difficulties and partial solutions to them). Some sources of detection error can be minimised with appropriate sampling effort ($\delta$ will converge to one with increasing number of samples), but other sources are often difficult to reduce (e.g. the occurrence of cryptic species might require molecular analysis for appropriate taxonomic identification as in~\citealt{Wirta2014,Frost2016}).

@@

    The considerations above raise a major challenge: when faced with empirical data, how may we infer whether unobserved interactions went undetected due to sampling or whether they truly do not occur? How then may we refine our sampling approaches to reduce uncertainties, and do we gain insights into the impact of multiple processes on field observations? Importantly, some sources of uncertainty can be minimised with appropriate sampling design and efforts while other sources are difficult or impossible to reduce since they are generated by chance variation created by the very process in which we are interested.

@@

          As a fourth and final approach to obtain the prior distribution of $\lambda$, we envisage using the outcome of a trait-matching model, provided such a model has been documented using external data and relevant traits are available. 
          In such as case, the prior distribution would follow the function $P(\lambda|\mathbf{T})=f(\mathbf{T})$ based on a set of traits for both species $\mathbf{T}$. There are several techniques available to perform this inference of interaction probability, some of which are Bayesian, and we refer to ~\citep{Bartomeus2016} and~\citep{Weinstein2017} for recent reviews about this topic. Note that in this case the prior might not be beta distributed and numerical methods might be required to compute the posterior distribution.
